---
title: "Introduction to cancR"
author: "Mathias Ørholt"
output:
  html_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an introduction to the cancR package.The article describes the workflow and main functions in the cancR package with in-depth examples. 

# Preparation

The cancR package is installed through github, see separate article.

```{r message = FALSE, warning = FALSE}
library(cancR)
```

To demonstrate the functions useful in data management, we load the simulated redcap dataframe

```{r}
df <- redcap_df
```

# Data management

The following functions are useful in data management tasks

## summarisR

To get a graphical glimpse of the data we can use the summarisR() function:

```{r}
#summarisR(df, 
       #   vars = c(sex, size, type, localisation, necrosis, cd10, sox10, ck))
```

## missR
And to exploit the number of missing values we use the missR() function

```{r}
missR(df)
```


We can also check if numerical variables are normally distributed with the distributR() function

## distributR

```{r}
distributR(df,
           vars = size)
```

