---
title: "Data Management"
output:
  html_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article is a thorough guide to data management in our. It mainly uses the dplyr package with additional cancR functions. The article is structured as chapters, where each chapter describes a specific task and should work as a library.

## Loading the cancR package

First we load the cancR package. The package automatically loads many packages useful for data management

```{r, message = FALSE, warning = FALSE}
library(cancR)
```

## Data

The cancR package comes with ready-to-use datasets. In this article we use the redcap_df dataset which imitates a dataset exported directly from redcap.

## Combining functions with the piping operator

It is advised to combine the functions described in this article into one code chunk that runs all functions at once. The functions are combined with the symbol `%>%` called a pipe. The shortcut for a pipe is ctrl+shift+m.

Piping starts by specifying the dataset of which the analyses should be performed. After this all subsequent functions are separated by a pipe. In the following example we start in the dataset "redcap_df", where we subsequently select the variables id, sex and birth, add a new variable called "new_variable" and lastly filter so that we subset the dataset to rows where type = 1. All these functions are combined into a piping structure and assigned to a new object named "new_data"

```{r}
new_data <- redcap_df %>% 
  select(id, sex, birth, type) %>% 
  mutate(new_variable = "new") %>% 
  filter(type == 1)
```



# Data inspection

Before starting on data management, it is important to get an overview of the dataset.

The entire dataset can be expected by either clicking on the dataset in the environment below or with the command `View()`

For a simpler inspection, `head()` shows only the first six rows in the console

```{r}
head(redcap_df)
```

All column names can be shown with `names()`

```{r}
names(redcap_df)
```


It is also important to assess the structure of the data to check for correct formatting. E.g. are date-variables coded as dates, continouous variables as numeric etc.

```{r}
str(redcap_df)
```

Here we see that all date variables are coded as characters and not date. The conversion to date are described in the chapter: "Date formatting"

To get a graphical glimpse of the data we can use the summarisR() function:

```{r, message = FALSE, warning = FALSE}
summarisR(redcap_df, 
          vars = c(sex, size, type, localisation, necrosis, cd10, sox10, ck))
```

And to exploit the number of missing values we use the missR() function

```{r}
missR(redcap_df)
```

We can also check if numerical variables are normally distributed with the distributR() function

```{r}
distributR(redcap_df,
           vars = size)
```

# Data management

The next section goes through the most basic data management functions from the dplyr package.

## Selection of variables

Variables/columns can be selected and removed with the select() function. 

```{r}
redcap_df %>% 
  select(id, sex, birth) %>% 
  head
```

Variables are removed with a minus sign.

```{r}
redcap_df %>% 
  select(-id, -birth) %>% 
  head
```

It is also possible to choose variable based on text patterns, which is useful for variables with a common prefix/suffix such as _date

```{r}
redcap_df %>% 
  select(contains("_date")) %>% 
  head
```

The text pattern can also be starts_with, ends_with and matches for an exact match.

If we need to select a large range of variables we call the first and last separated by a colon:

```{r}
redcap_df %>% 
  select(sex:sox10) %>% 
  head
```


## Renaming variables

Renaming of variable names can be done using rename(). The syntax is "new name" = "old name"

```{r}
redcap_df %>% 
  rename(index = date_of_surgery,
         cytokeratin = ck) %>% 
  head
```

## Create/modify variables

Variables can be created or modified with the mutate() function with the syntax: mutate(variable = condition). If the variable already exists in the dataset, it is modified automatically.

We now recode necrosis, so that 1 = yes and everything else is "no".

```{r}
redcap_df %>% 
  mutate(necrosis = ifelse(necrosis == 1, "yes", "no")) %>% 
  head
```

Notice that missing values in necrosis remain missing. If we want to also assign these as "no" we change the `==` in the mutate function to `%in%`. This will also imply that missing values are converted to the "else" statement, here "no".

```{r}
redcap_df %>% 
  mutate(necrosis = ifelse(necrosis %in% 1, "yes", "no")) %>% 
  head
```


If we want more explicit control with the recoding or we have more than one condition, we use case_when

```{r}
redcap_df %>% 
  mutate(necrosis = case_when(necrosis == 1 ~ "yes",
                              necrosis == 0 ~ "no")) %>% 
  head
```

Now we have preserved the missing values. We can also control what to do with values that does not satisfy any of the criteria

```{r}
redcap_df %>% 
  mutate(necrosis = case_when(necrosis %in% 1 ~ "yes",
                              necrosis %in% 0 ~ "no",
                              T ~ "missing")) %>% 
  head
```

### Collecting multiple mutate functions

Multiple mutate functions can be collected in one call

```{r}
redcap_df %>% 
  mutate(new_variable = "new",
         sex = ifelse(sex == 1, "f", "m"),
         size = case_when(size > 40 ~ "large",
                          size < 10 ~ "small",
                          T ~ "intermediate")) %>% 
  head
```

### Mutating multiple variables simultaneously

Multiple variables can be modified with `across()` within mutate. The syntax is:\

`mutate(across(c(variable1, variable2), ~ function))

Here we convert the variables cd10, sox10 and ck to characters

```{r}
redcap_df %>% 
  select(cd10, sox10, ck) %>% 
  mutate(across(c(cd10, sox10, ck), ~ as.character(.))) %>% 
  head
```

The `.` inside the `as.character(.)` refers to all the variables inside `across()`. 

## Recoding of variables

Recoding of variables can be done with the recodR() function in the cancR package. The syntax here is a list of lists, so that list(variable = list(new name = old name))

```{r}
redcap_df %>% 
  recodR(list("sex" = 
                list("female" = 1,
                           "male" = 2),
              "type" = 
                list("benign" = 0,
                            "in_situ" = 1,
                            "malignant" = 2),
              "localisation" = 
                list("head" = 1,
                     "neck" = 2,
                     "trunk" = 3,
                     "upper_extremity" = 4,
                     "lower_extremity" = 5))) %>% 
  head
```

If the recoding should be more advanced and should be based on one or more conditions, we can use ifelse() or case_when() (see examples under "Create/modify variables").

## Date formatting

Dates are often formatted as character strings and need to be converted to correct date format. This can easily be done with the datR() function:

```{r}
redcap_df %>% 
  datR(c(contains("date"), birth, followup)) %>% 
  head
```

## Categorization of continuous variables

The optimal method for splitting continuous variables depends on the number of splits:
- One splits: ifelse()
- More than one split: case_when()
- Splits based on a sequence or quantiles: cutR()

One split with ifelse()

```{r}
redcap_df %>% 
  mutate(size_bin = ifelse(size > 20, "large", "small")) %>% 
  head
```

More splits with case_when()

```{r}
redcap_df %>% 
  mutate(size_bin = case_when(size > 40 ~ "large",
                          size < 10 ~ "small",
                          T ~ "intermediate")) %>% 
  head
```

Splits based on a sequence can be done with the cutR() function:

```{r}
redcap_df %>% 
  cutR(size, 
       seq(0,50,10)) %>% 
  head
```

Multiple splits can also be performed with cutR() with name assigning

```{r}
redcap_df %>%
  cutR(vars = c(age, size),
       seqlist = list("age" = seq(0,100,10),
                      "size" = list("quantile", c(0,0.25,0.5,0.75,1))),
       name.list = list("age" = "age_group",
                        "size" = "size_bin")) %>% 
  head

```

The new variables can also be given the same name pattern if the input variables are similar such as dates

```{r}
redcap_df %>% 
  #Conversion into date format
  datR(contains("date")) %>% 
  cutR(vars = c(recurrence_date, metastasis_date),
       seqlist = list("recurrence_date" = seq(1900,2030,10),
                      "metastasis_date" = seq(1900,2030,10)),
       name.pattern = "_bin")
```

## Conversion to factor

In many cases it is of interest to specify a categorical variable with certain levels or reference groups. This is done with factR(). If nothing else is specified, factR() converts a character string into a factor, with the level with most observations as reference. The function str() is used to show the formatting of the type variable has changed to factor.

```{r}
redcap_df %>% 
  factR(type) %>% 
  str
```

The reference group is specified using the reference argument

```{r}
redcap_df %>% 
  factR(type,
        reference = "0") %>% 
  str
```

Levels can be manually assigned

```{r}
redcap_df %>% 
  factR(type,
        levels = c("2","1","0")) %>% 
  str
```

New labels can also be assigned and automatically specify levels simultaneously

```{r}
redcap_df %>% 
  factR(type,
        labels = list("type" = list("benign" = "0",
                                    "intermediate" = "1",
                                    "malignant" = "2")),
        lab_to_lev = T) %>% 
  str
```

Lastly, all the arguments can be specified for multiple variables at once

```{r}
redcap_df %>% 
  factR(vars = c(type, sex),
        reference = list("sex" = "2"),
        levels = list("type" = c("0", "1", "2")),
        labels = list("sex" = c("f" = "1",
                                "m" = "2"),
                      "type" = c("benign" = "0",
                      "intermediate" = "1",
                      "malignant" = "2"))) %>% 
  str
```

## Subset rows (filters)

If we want to keep only certain rows we use filter(). Here we limit the dataset to patients without necrosis (necrosis = 0)

```{r}
redcap_df %>% 
  filter(necrosis == 0) %>% 
  head
```

In case of multiple conditions we use the `%in%` operator, as this specified that the variables has one of the following values

```{r}
redcap_df %>% 
  filter(localisation %in% c(1,2,3)) %>% 
  head
```


For numerical variables we use <, >, ==, >= and <=

```{r}
redcap_df %>% 
  filter(size > 20) %>% 
  head
```

We can also use between() to specify an interval

```{r}
redcap_df %>% 
  filter(between(size, 10,20)) %>% 
  head
```

Multiple conditions can be combined using so-called "boolean" operators whic are and/or/not etc. Here we keep rows where necrosis = 1 AND localisation is 1, 2 or 3 OR size is larger than 10. Here necrosis needs to be 1, but either of the conditions for localisation or size can be satisified.

```{r}
redcap_df %>% 
  filter(necrosis == 1 & (localisation %in% c(1,2,3) | size > 10)) %>% 
  head
```

## Arranging/sorting data

Sorting is done with the arrange(). If the variable needs to be in descending order it is prefixed by a desc()

```{r}
redcap_df %>% 
  arrange(size) %>% 
  head()
```

Descending size

```{r}
redcap_df %>% 
  arrange(desc(size)) %>% 
  head
```


## Missing data

To get a quick overview of missing data in the dataset we use missR()

```{r}
redcap_df %>% 
  missR()
```

Missing values can be dropped with drop_na()

```{r}
redcap_df %>% 
  drop_na(metastasis_date, necrosis) %>% 
  head
```

Sometimes we want to remove rows where at least one of the values are NA. This is done with rowR() which is useful for rowwise operations

```{r}
redcap_df %>% 
  rowR(vars = c(cd10, sox10, ck),
       type = "any.na",
       filter = "remove") %>% 
  head
```

We can also flag the rows into a new variable

```{r}
redcap_df %>% 
  rowR(vars = c(cd10, sox10, ck),
       type = "any.na",
       new = "flag") %>% 
  head
```

## Join multiple data frames

The terminology is that we have a dataset X and a dataset Y.There are different types of joins, as joins can be side-by-side (such as merging to datasets based on a key variable) and end-to-end where one dataside is a continuation of another. In the side-by-side joins we need to specify a "by/key" variable which specifies how we link the two datasets. In the end-to-end join there is never a "key" as these are extensions and should roughly have the same variables.

### Left join
In a left_join, dataset X is complete and only observations from dataset Y that are present in X are joined. For all

If we want to join two data frames, we use left_join().
```{r}
left_join(redcap_df, analysis_df, by = "id") %>% 
  tail
```

We see that some of the patients do not have any rows, as the redcap_df only has ids up to 350.

We can left_join in a pipe. Here the current dataset we are working with is symbolised by a dot. We can also use other functions with the pipe operator inside the left_join.

```{r}
analysis_df %>% 
  arrange(id) %>% 
  left_join(.,
            redcap_df %>% select(id, contains("date")),
            by = "id") %>% 
  tail
```

### Full join
In a full join, all rows are combined

```{r}
analysis_df %>% 
  full_join(.,
            redcap_df %>% select(id, contains("date")),
            by = "id") %>% 
  tail
```

Here we se missing data for dataset X and complete data for dataset Y as all rows are kept.

### Appending rows (bind rows)

To obtain a relevant case we split our dataset into two

```{r}
r1 <- redcap_df %>% filter(id < 100)
r2 <- redcap_df %>% filter(id >= 100)

#Binding it back together
bind_rows(r1, r2) %>% head
```

