---
title: "Import data to R"
output:
  html_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message = FALSE, warning = FALSE}
library(cancR)
```

All data files should optimally be located in a `data` folder in the project folder. If the project is set up correctly (see the article `R setup`) we can call the files directly from the `data` folder: "data/file.csv".

# Loading data
The `readR()` function in `cancR` makes loading easy by automatically chosing the correct function based on the file extension.

```{r, eval = F}
mydata <- readR("data/dataset.csv")
```

`readR()` currently handles the extensions: `csv`, `xlsx`, `rds` and `txt`.


# Loading data from redcap

Data from redcap can conveniently be loaded with `redcapR()`. This function takes two arguments: 1) The dataset exported from redcap (see how to export from redcap in the article `Redcap and R`) and 2) The data dictionary from redcap (see `Redcap and R`)

```{r, eval = F}
redcap_data <- readR("data/redcap_data.csv")
dictionary <- readR("data/data_dictionary.csv")

mydata <- redcapR(redcap_data,
                  dictionary)
```

This will apply all labels from redcap automatically so that manual recoding is not necessary.

If there is a sheet containing CPR-numbers and id numbers it is also possible to automatically extract birth date and sex in `redcapR`

```{r, eval = F}
redcap_data <- readR("data/redcap_data.csv")
dictionary <- readR("data/data_dictionary.csv")
cprlist <- readR("data/cprlist.csv")

mydata <- redcapR(redcap_data,
                  dictionary,
                  cprlist)
```

