---
title: "Statistical Analysis Plan"
output:
  html_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The statistical analysis plan ensures an a priori defined analysis plan, transparency for all collaborators and a optimal workflow.

The statistical analysis plan should contain the following:

## Authors
List of the full author list or at least tentative authors

## Purpose
Short introduction to the project

## Population
Describe the population

## Inclusion/Exclusion criteria
Specify the above criteria in bullet points

## Exposure
Define the exposure including the type of data (continuous, binary, categorical, ordinal)

## Outcome
Define the outcome including type of data (continuous, binary, categorical, ordinal, time-to-event)

## Covariates
Specify if any covariates should be accounting for, preferably with a directed acyclic graph (DAG)

## Clustering
Specify if there are clustered observations, if yes specify the structure (e.g. breasts within patients, samples within patients) 

## Analysis hierarchy
Specify the order of analyses. Specify the role of each analysis (primary, secondary, subgroup, sensitivity)\
\
- Primary analysis: Risk of all complications (hematoma, explantation and infection) at 5 years\
- Secondary analysis: Same as above for each of the complications\
- Subgroup: Same as the primary analysis but for patients between 2018-2024\

## Data dictionary
A table with four columns specifying\
\
1) The variable name in the dataset
2) Very short description of the variable
3) type of data.\
- If continuous specify the distribution (0 to 1, non-negative, exponential)\
- If categorical specify the levels and if there is any ordering such as (low, moderate, severe)\
4) The use of the data point (e.g. outcome, exposure, covariate, cluster)

```{r, echo = F}
flextable::flextable(data.frame(name = c("hema", "t_hema", "imp_plane2", "age"),
           description = c("whether hematoma occured",
                          "time to hematoma",
                          "implant plane",
                          "age at surgery"),
           type = c("binary, 0 = censuring, 1=event",
                    "continuous, non-negative",
                    "binary, sg/sm",
                    "ordinal, binned 5-year age groups"),
           role = c("outcome", "outcome", "exposure", "covariate")), cwidth = 2)
```

## Path
The path to the project folder

# R setup
The main working folder should ONLY contain the following files at the beginning of the project.\
\
- Project file (.Rproj)\
- MAIN_SCRIPT (.R)\
- Statistical analysis plan (.docx)\
- data folder with ONE dataset (.csv/.rds/.txt/.parquet)

See the article `Setup R` for help.
